A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-03-16 paper/paper.tex
Adding magic compilation incantation for hal/arxiv
paper: fix some problems pointed out by reviewers

\documentclass[10pt]{art.cls/art}

@@

\begin{document}
\maketitle
\begin{abstract}
  Constructions such as semi-simplicial and semi-cubical sets can be defined in the ``usual way'' as presheaves over respectively, the semi-simplex or semi-cube category, which we call \emph{fibered} definitions, but also defined like in e.g. Voevodsky~\cite{voevodsky12} or in previous work~\cite{herbelin15}, as a dependently-typed construction, which we call \emph{indexed}. %TODO anonymize

@@

\section{Introduction}
In the context of functional programming, Reynolds' parametricity~\cite{reynolds72} interprets types as relations characterizing the observational behavior of programs of this type. Parametricity can be iterated, and it has been noted that iterated Reynolds' parametricity, when using more generally graphs than relations, has a cubical flavor~\cite{johann17,altenkirch15,moulin16,moeneclaey21,moeneclaey22phd}. We obtain a \emph{unary} variant of Reynolds' \emph{binary} parametricity by using predicates or families instead of relations or graphs, in which case, we obtain a form of realizability~\cite{bernardy12,lasson12,moulin16}. It has then been noted that iterated unary parametricity has an augmented simplicial flavor~\footnote{Private communication with Hugo Moeneclaey and Thorsten Altenkirch}. % TODO anonymize

We exploit the connection between iterated unary parametricity and augmented simplicial sets, and between iterated binary parametricity and cubical sets to present a uniform construction of both augmented semi-simplicial and semi-cubical sets obtained by applying simple rules characterizing iterated parametricity. In contrast to the standard presheaf presentation of augmented semi-simplicial and semi-cubical sets~\cite{fri08,grandis03,buchholtz17}, our uniform construction, which we call \emph{$\nu$-sets}, is \emph{indexed}. That is, instead of having the set of augmented semi-simplices or semi-cubes in dimension $n+1$ \emph{fibered} over the set of augmented semi-simplices or semi-cubes in dimension $n$, we consider families of augmented semi-simplices and semi-cubes indexed over their faces.

The outline of the paper is as follows. First, we recall the definition of the augmented semi-simplicial and semi-cubical categories using a combinatorial presentation, that directly leads to our generalization to $\nu$-sets. We then proceed to explain our indexed construction, with the help of diagrams. Before spelling out the details of our formalization in type-theoretic language, we have a section on reasoning with sets inside type theory. We end with some of the finer details of our mechanization.

Our precise mechanization in Coq is available at \href{https://github.com/artagnon/bonak}{github.com/artagnon/bonak}. The construction was conceived in Summer 2019, and the mechanization began in late 2019. A sketch of the construction was presented at the 2020 HoTT-UF workshop and the completion of the mechanization was reported at the TYPES 2022 conference.

\section{Semi-simplicial and semi-cubical sets}
\subsection{Augmented semi-simplicial sets}
Augmented semi-simplicial sets are defined similarly to semi-simplicial sets, except that the connected components are additionally dependent on a ``color''. Conversely, semi-simplicial sets can be seen as augmented semi-simplicial sets over a singleton set of colors.

@@

\begin{notation}[Finite sequences]
  We denote finite sequences by $\{i_1, \ldots i_n\}$ for $i_j$ ranging over some domain. In particular, the empty sequence is written $\{\}$ and we define $i :: \{j_1, \ldots, j_n\}$ to be $\{i, j_1, \ldots, j_n\}$.
\end{notation}

\begin{definition}[$\DeltaPlus$]
  \begin{figure*}[!t]
    \begin{align*}
      \Obj(\DeltaPlus) & := \Nat                                                                     \\
      \Hom(p, n)       & := \{l \in \{0, \kstar\}^n \mid \text{number of $\kstar$ in $l = p$}\}      \\
      g \circ f        & :=
      \begin{cases}
        f                  & \text{if $g = \{\}$}                                                      \\
        0 :: (g' \circ f)  & \text{if $g = (0 :: g')$}                                                 \\
        a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = 0$ or $\kstar$} \\
      \end{cases} \\
      \id              & := \{\kstar, \ldots, \kstar\} \text{ $n$ times for $\id \in \Hom(n, n)$}
    \end{align*}
    \caption{Definition of \DeltaPlus}\label{fig:deltaplus}
  \end{figure*}

@@

\begin{example}[$\DeltaPlus^2$]
  In dimension $2$, the standard augmented semi-simplex is drawn as two points, given by $\Hom(1, 2)$, along with a line connecting them, given by $\Hom(2, 2)$. We use the color black to denote the unique morphisms in $\Hom(0, 1)$ and $\Hom(0, 2)$.
  \begin{equation*}
    \begin{tikzcd}
      \kstar0 \arrow[r, dash, "\kstar\kstar"] & 0\kstar
    \end{tikzcd}
  \end{equation*}
\end{example}

@@

\begin{definition}[\Cube]
  The definition of $\Cube$ is shown in Fig.~\ref{fig:cube}. The symbols $L$ and $R$ indicate opposite faces of a cube.
  \begin{figure*}[!t]
    \begin{align*}
      \Obj(\Cube) & := \Nat                                                                                \\
      \Hom(p, n)  & := \{l \in \{L, R, \kstar\}^n \mid \text{number of $\kstar$ in $l = p$}\}              \\
      g \circ f   & :=
      \begin{cases}
        f                  & \text{if $g = \{\}$}                                                            \\
        a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $a = L$ or $R$}                          \\
        a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = L$, $R$, or $\kstar$} \\
      \end{cases} \\
      \id         & := \{\kstar, \ldots, \kstar\} \text{ $n$ times}
    \end{align*}
    \caption{Definition of \Cube}\label{fig:cube}
  \end{figure*}

@@

A $\nu$-set is thus a functor $\phi$ from the $\nu$-semi-shape category to $\Set$ and we call $n$-$\nu$-semi-shape an element of $\phi(n)$. As in the augmented semi-simplicial and semi-cubical cases, the standard $(n + 1)$-$\nu$-semi-shape is obtained by connecting together $\nu$ copies of the standard $n$-$\nu$-semi-shape with an extra copy stretched in the new dimension. We clarify in the next sections how this process of construction is similar to the parametricity translation developed for functional programming~\cite{reynolds72} and more generally for type theory~\cite{bernardy10, bernardy11, atkey14, bernardy15}.

\section{Our setting}
We now explain how we recover the notion of sets in type theory, explain the difference between the fibered and indexed representations, and supply an example of indexed semi-cubical sets.

\subsection{Working with sets in type theory}
Martin-Löf's Type theory~\cite{martinlof75,martinlof84} is a logical formalism based on the notion of a \emph{type} rather than that of a \emph{set}. It can be seen as a foundation of mathematics alternative to set theory and is the core of several tools for the formalization of mathematics such as Agda~\cite{agda23}, Coq~\cite{coq22} or Lean~\cite{lean15}. In type theory, propositions are types and proofs are programs. A particularity of type theory is also that types and programs, henceforth propositions and proofs also, are considered modulo an equational theory called definitional equality.

Type theory is a flexible formalism supporting different models. Some models are based on topological spaces, where equality is interpreted as path, and substitutivity of equality as transport. This model supports the univalence principle stating that equality of types mimics equivalence of types, leading to the development of Homotopy Type Theory~\cite{hottbook}. In type theory, types are organized in a hierarchy of universes written $\Type_m$ for $m$ a natural number. Main types in type theory are the types of dependent pairs, written $\Sigma a : A.\,B(a)$, the types of dependent functions, written $\Pi a:A.\,B(a)$, for $A$ a type and $B(a)$ a type dependent on the inhabitant $a$ of $A$, and the type of propositional equality, written $t = u$. We assume our type theory to also include a distinguished singleton type, written $\unittype$, and with inhabitant $\unitpoint$, the type of boolean values, called $\textsf{bool}$, and the type of natural numbers. We also write $\hd$ and $\tl$ the projections of dependent pairs, and $\refl$ for reflexivity. Logical propositions being types themselves, we use $\Pi$ to represent universal quantification and $\Sigma$ to represent existential quantification.

@@

\subsection{Fibered versus indexed representation}
It is well-known in set theory that families of small sets can be represented in a fibered or indexed way. In the fibered way, a family over $T$ is represented as a pair of a set $S$, and a function from $S$ to $T$. The function classifies the original set into fibers. In contrast, in the indexed way, a family over $T$ is represented as a functional relation with domain $T$, which itself, by the replacement axiom, is a set.

In category theory, the correspondence between the fibered and indexed way manifests as the correspondence between Grothendieck fibrations and pseudofunctors, of which a particular instance is the correspondence between discrete fibrations and presheaves, or even between discrete fibrations over sets and families of sets.

The latter correspondence can be expressed in Homotopy Type Theory, for $T: \Type$, as an equivalence between $\Sigma S: \Type.\; S \rightarrow T$ and $T \rightarrow \Type$, which can be proved using univalence. This correspondence also holds on \U, for $T: \U$:
\begin{equation*}
  \Sigma S: \U.\, S\rightarrow T ~\simeq~ T \rightarrow \U
\end{equation*}

\subsection{Example: indexed semi-cubical sets}
Let us now come back to semi-cubical sets. A semi-cubical set can be represented as a sequence of fibrations, together with appropriate coherence conditions.
\begin{equation*}
  \begin{tikzcd}
    X_0: \U & X_1: \U \arrow[l, "\delta^L" description, shift left=2] \arrow[l, "\delta^R" description, shift right=2] & X_2: \U \arrow[l, "\delta^{L\kstar}" description, shift left=6] \arrow[l, "\delta^{R\kstar}" description, shift left=2] \arrow[l, "\delta^{\kstar R}" description, shift right=2] \arrow[l, "\delta^{\kstar L}" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}

@@

Now, one may wonder how to express the type of $X_n$ uniformly as a function of $n$. Connecting semi-cubical sets to binary parametricity will give us a solution which we describe in the next section.

\subsection{Relating the indexed representation to parametricity}
The process of construction of the type of $X_1$ from that of $X_0$, and from the type of $X_2$ to that of $X_1$ in the last section, is similar to applying a binary parametricity translation and expecting the resulting translation to be inhabited. The parametricity which we consider interprets a type $A$ by a graph $A_\kstar$ over this type, and a term $t: A$ as an edge in $A_\kstar(t)(t)$. In particular, $\U$ is interpreted as the graph ${\U}_\kstar$, which takes $A_L$ and $A_R$ in $\U$ and returns the type $A_L \rightarrow A_R \rightarrow \U$ of graphs over $A_L$ and $A_R$. Also, for $A$ interpreted by $A_\kstar$ and $B$ interpreted by $B_\kstar$, a dependent function type $\Pi a: A.\, B$ is interpreted as the graph $(\Pi a: A.\, B)_\kstar$ that takes two functions $f_L$ and $f_R$ of type $\Pi a: A.\, B$, and expresses that these functions map related arguments in $A$ to related arguments in $B$:
\begin{align*}
   & (\Pi a: A.\, B)_\kstar(f_L)(f_R) \defeq                                                      \\
   & \Pi a_L: A.\, \Pi a_R: A.\, \Pi a_\kstar: A_\kstar(a_L)(a_R).\, B_\kstar(f_L(a_L))(f_R(a_R))
\end{align*}

@@

\subsection{Intuition for our formal construction}
To assign types to $X_0, X_1, X_2, \ldots$ in the above indexed representation, we give a recursive definition relying on building blocks, which we call $\framep$, $\layer$, and $\painting$. A $\framep$ is a boundary of a standard form (simplex, cube, etc.), which we decompose into $\layer$, and a $\painting$ corresponds to a filled $\framep$. Some $\framep$ are full and we call them $\fullframe$.

>>>>

%% -*- latex-command: pdflatex -shell-escape -*-
\documentclass[10pt]{art.cls/art}

@@

\begin{document}
\maketitle
\begin{abstract}
  Constructions such as semi-simplicial and semi-cubical sets can be defined in the ``usual way'' as presheaves over respectively, the semi-simplex or semi-cube category, which we call \emph{fibered} definitions, but also defined like in e.g. Voevodsky~\cite{voevodsky12} or in previous work~\cite{herbelin15}, as a dependently-typed construction, which we call \emph{indexed}.

@@

\section{Introduction}
In the context of functional programming, Reynolds' parametricity~\cite{reynolds72} interprets types as relations characterizing the observational behavior of programs of this type. Parametricity can be iterated, and it has been noted that iterated Reynolds' parametricity can be interpreted in various various cubical sets~\cite{johann17,altenkirch15,moulin16,moeneclaey21,moeneclaey22phd}. We obtain a \emph{unary} variant of Reynolds' \emph{binary} parametricity by using predicates or families instead of relations or graphs, in which case, we obtain a form of realizability~\cite{bernardy12,lasson12,moulin16}. It has then been noted that iterated unary parametricity has an augmented simplicial flavor~\footnote{Private communication with Hugo Moeneclaey and Thorsten Altenkirch}.

We exploit the connection between iterated unary parametricity and augmented simplicial sets, and between iterated binary parametricity and cubical sets to present a uniform construction of both augmented semi-simplicial and semi-cubical sets obtained by applying rules inspired from iterated parametricity. In contrast to the standard presheaf presentation of augmented semi-simplicial and semi-cubical sets~\cite{fri08,grandis03,buchholtz17}, our uniform construction, which we call \emph{$\nu$-sets}, is \emph{indexed}. That is, instead of having the set of augmented semi-simplices or semi-cubes in dimension $n+1$ \emph{fibered} over the set of augmented semi-simplices or semi-cubes in dimension $n$, we consider families of augmented semi-simplices and semi-cubes indexed over their faces.

The outline of the paper is as follows. We recall the definition of the augmented semi-simplicial and semi-cubical categories using a combinatorial presentation, that directly leads to our generalization to $\nu$-sets, in \ref{sec:nu}. We then proceed to explain our setting in \ref{sec:setting}. Before spelling out the details of our formalization in type-theoretic language in \ref{sec:itt}, we explain the connection between the indexed representation and parametricity in \ref{sec:indexed-param}, and build up intuitions in \ref{sec:intuition}. We end with some of the finer details of our mechanization in \ref{sec:mech}.

See \href{https://github.com/artagnon/bonak}{github.com/artagnon/bonak} for our mechanization. The construction was conceived in Summer 2019, and the mechanization began in late 2019. A sketch of the construction was presented at the 2020 HoTT-UF workshop and the completion of the mechanization was reported at the TYPES 2022 conference.

\section{Semi-simplicial and semi-cubical sets\label{sec:nu}}
\subsection{Augmented semi-simplicial sets}
Augmented semi-simplicial sets are defined similarly to semi-simplicial sets, except that the connected components are additionally dependent on a ``color''. Conversely, semi-simplicial sets can be seen as augmented semi-simplicial sets over a singleton set of colors.

@@

\begin{notation}[Finite sequences]
  We denote finite sequences by $[i_1, \ldots i_n]$ for $i_j$ ranging over some domain. In particular, the empty sequence is written $[]$ and we define $i :: [j_1, \ldots, j_n]$ to be $[i, j_1, \ldots, j_n]$.
\end{notation}

\begin{definition}[$\DeltaPlus$]
  \begin{figure*}[!t]
    \begin{align*}
      \Obj(\DeltaPlus) & := \Nat                                                                     \\
      \Hom(p, n)       & := \{l \in [0, \kstar]^n \mid \text{number of $\kstar$ in $l = p$}\}        \\
      g \circ f        & :=
      \begin{cases}
        f                  & \text{if $g = []$}                                                        \\
        0 :: (g' \circ f)  & \text{if $g = (0 :: g')$}                                                 \\
        a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = 0$ or $\kstar$} \\
      \end{cases} \\
      \id              & := [\kstar, \ldots, \kstar] \text{ $n$ times for $\id \in \Hom(n, n)$}
    \end{align*}
    \caption{Definition of \DeltaPlus}\label{fig:deltaplus}
  \end{figure*}

@@

\begin{example}[$\DeltaPlus^2$]
  The standard augmented $2$-semi-simplex is drawn as two points, given by $\Hom(1, 2)$, along with a line connecting them, given by $\Hom(2, 2)$. We use the color black to denote the unique morphisms in $\Hom(0, 1)$ and $\Hom(0, 2)$.
  \begin{equation*}
    \begin{tikzcd}
      \kstar0 \arrow[r, dash, "\kstar\kstar"] & 0\kstar
    \end{tikzcd}
  \end{equation*}
\end{example}

@@

\begin{definition}[\Cube]
  The definition of $\Cube$ is shown in Fig.~\ref{fig:cube}. The symbols $L$ and $R$ indicate opposite faces of a cube.
  \begin{figure*}[!t]
    \begin{align*}
      \Obj(\Cube) & := \Nat                                                                                \\
      \Hom(p, n)  & := \{l \in [L, R, \kstar]^n \mid \text{number of $\kstar$ in $l = p$}\}                \\
      g \circ f   & :=
      \begin{cases}
        f                  & \text{if $g = []$}                                                              \\
        a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $a = L$ or $R$}                          \\
        a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = L$, $R$, or $\kstar$} \\
      \end{cases} \\
      \id         & := [\kstar, \ldots, \kstar] \text{ $n$ times}
    \end{align*}
    \caption{Definition of \Cube}\label{fig:cube}
  \end{figure*}

@@

A $\nu$-set is thus a contravariant functor $\phi$ from the $\nu$-semi-shape category to $\Set$ and we call $n$-$\nu$-semi-shape an element of $\phi(n)$. As in the augmented semi-simplicial and semi-cubical cases, the standard $(n + 1)$-$\nu$-semi-shape is obtained by connecting together $\nu$ copies of the standard $n$-$\nu$-semi-shape with an extra copy stretched in the new dimension. We clarify in the next sections how this process of construction is similar to the parametricity translation developed for functional programming~\cite{reynolds72} and more generally for type theory~\cite{bernardy10, bernardy11, atkey14, bernardy15}.

\section{Our setting\label{sec:setting}}
We now explain how we recover the notion of sets in type theory, explain the difference between the fibered and indexed representations, and supply an example of indexed semi-cubical sets.

\subsection{Working with sets in type theory}
Martin-Löf's Type theory~\cite{martinlof75,martinlof84} is a logical formalism based on the notion of a \emph{type} rather than that of a \emph{set}. It can be seen as a foundation of mathematics alternative to set theory and is the core of several tools for the formalization of mathematics such as Agda~\cite{agda23}, Coq~\cite{coq22} or Lean~\cite{lean15}. In type theory, propositions are types and proofs are programs. A particularity of type theory is also that types and programs, hence propositions and proofs also, are considered modulo an equational theory called definitional equality.

Type theory is a flexible formalism supporting different models. Some models are based on topological spaces, where equality is interpreted as path, and substitutivity of equality as transport. These models support the univalence principle stating that equality of types mimics equivalence of types, leading to the development of Homotopy Type Theory~\cite{hottbook}. In type theory, types are organized in a hierarchy of universes written $\Type_m$ for $m$ a natural number. Main types in type theory are the types of dependent pairs, written $\Sigma a : A.\,B(a)$, the types of dependent functions, written $\Pi a:A.\,B(a)$, for $A$ a type and $B(a)$ a type dependent on the inhabitant $a$ of $A$, and the type of propositional equality, written $t = u$. We assume our type theory to also include a distinguished singleton type, written $\unittype$, and with inhabitant $\unitpoint$, the type of boolean values, called $\textsf{bool}$, and the type of natural numbers. We also write $\hd$ and $\tl$ the projections of dependent pairs, and $\refl$ for reflexivity. Logical propositions being types themselves, we use $\Pi$ to represent universal quantification and $\Sigma$ to represent existential quantification.

@@

\subsection{Relating the indexed representation to parametricity\label{sec:indexed-param}}
% \begin{equation*}
%   \Sigma S: \U.\, S\rightarrow T ~\simeq~ T \rightarrow \U
% \end{equation*}

Let us now come back to semi-cubical sets. A semi-cubical set can be represented as a sequence of fibrations, together with appropriate coherence conditions.
\begin{equation*}
  \begin{tikzcd}
    X_0: \U & X_1: \U \arrow[l, "\delta^L" description, shift left=2] \arrow[l, "\delta^R" description, shift right=2] & X_2: \U \arrow[l, "\delta^{L\kstar}" description, shift left=6] \arrow[l, "\delta^{R\kstar}" description, shift left=2] \arrow[l, "\delta^{\kstar R}" description, shift right=2] \arrow[l, "\delta^{\kstar L}" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}

@@

The process of construction of the type of $X_1$ from that of $X_0$, and from the type of $X_2$ to that of $X_1$ in the last section, is similar to applying a binary parametricity translation and expecting the resulting translation to be inhabited. The parametricity which we consider interprets a type $A$ by a graph $A_\kstar$ over this type, and a term $t: A$ as an edge in $A_\kstar(t)(t)$. In particular, $\U$ is interpreted as the graph ${\U}_\kstar$, which takes $A_L$ and $A_R$ in $\U$ and returns the type $A_L \rightarrow A_R \rightarrow \U$ of graphs over $A_L$ and $A_R$. Also, for $A$ interpreted by $A_\kstar$ and $B$ interpreted by $B_\kstar$, a dependent function type $\Pi a: A.\, B$ is interpreted as the graph $(\Pi a: A.\, B)_\kstar$ that takes two functions $f_L$ and $f_R$ of type $\Pi a: A.\, B$, and expresses that these functions map related arguments in $A$ to related arguments in $B$:
\begin{align*}
   & (\Pi a: A.\, B)_\kstar(f_L)(f_R) \defeq                                                      \\
   & \Pi a_L: A.\, \Pi a_R: A.\, \Pi a_\kstar: A_\kstar(a_L)(a_R).\, B_\kstar(f_L(a_L))(f_R(a_R))
\end{align*}

@@

\subsection{Intuition for our formal construction\label{sec:intuition}}
To assign types to $X_0, X_1, X_2, \ldots$ in the above indexed representation, we give a recursive definition relying on building blocks, which we call $\framep$, $\layer$, and $\painting$. A $\framep$ is a boundary of a standard form (simplex, cube, etc.), which we decompose into $\layer$, and a $\painting$ corresponds to a filled $\framep$. Some $\framep$ are full and we call them $\fullframe$.
